Defining Transcriptome using RNA Sequencing\\ of {\em Streptococcus agalactiae}

### 2012-09-12 package-source/sagalactiae.Rnw
Manual html.

\documentclass{article}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{color} 
\usepackage{setspace} 
\usepackage{rotating}
\usepackage{natbib}
\usepackage{xr}
\usepackage{url}
\usepackage{../../latex/sty/genres}
\externaldocument{supplementary}

@@

\makeatother
\pagestyle{myheadings}
\markright{Transcriptome of Streptococcus mutans}

@@

\title{Defining Transcriptome using RNA Sequencing\\
of {\em Streptococcus mutans} }

\author{Sang Chul Choi$^{1}$, Charles D.\ Danko$^{1}$, 
Melissa J.\ Hubisz$^{1}$, \\
Paulina Dmitrievna Pavinski Bitar$^{2}$, \\
Robert A.\ Burne$^{3}$
Michael J.\ Stanhope$^{2}$, and Adam Siepel$^{1}$}

@@

\begin{footnotesize}
\begin{center}
$^1$Department of Biological Statistics and Computational Biology,\\
Cornell University, Ithaca, NY 14853, USA
\\[1ex]
$^2$Department of Population Medicine and Diagnostic Sciences,\\
College of Veterinary Medicine, Cornell University, Ithaca, NY 14853, USA
\\[1ex]
$^3$Department of Oral Biology,\\
College of Dentistry, University of Florida, Gainesville, Florida 32610 USA
\\
\end{center}
\end{footnotesize}

@@

\begin{tabular}{lp{4.5in}}
{\bf Submission type:}& Research Article
\vspace{1ex}\\
{\bf Keywords:}&Carbon catabolite repression, RNA-seq, non-coding RNAs
\vspace{1ex}\\
{\bf Running Head:}&Transcriptome of {\em Streptococcus mutans}
\vspace{1ex}\\ 
{\bf Corresponding Author:}&
\begin{minipage}[t]{4in}
 Adam Siepel\\
 102E Weill Hall, Cornell University\\
 Ithaca, NY 14853\\
 Phone: +1-607-254-1157\\
 Fax: +1-607-255-4698\\
 Email: acs4@cornell.edu
\end{minipage}
\end{tabular}

@@

\setcounter{page}{2}

@@

\fixme{Add more specific findings while summarizing the results. Connect two
ideas of the differential expression study and transcriptome study.}

A bacterial transcriptome of the primary etiological agent of human dental
caries, \textit{Streptococcus mutans}, is described using deep RNA sequencing.
The technique of RNA-Seq has allowed 
the determination of transcript boundaries for annotated genes and
the identification of previously unknown transcribed regions 
including mRNAs and small noncoding RNAs in the genome.

Differential expression studies on the role of CcpA in 
carbon catabolite repression (CCR) suggested that some of the novel mRNAs and
small RNAs found in our study could be involved in the regulation of CcpA.

Our study provides more precise landscape of transcriptome changes in response to
different carbon sources in \textit{S. mutans}.

The knowledge that we gained from the current study of \textit{S. mutans}
transcriptome will lead to re-evaluation of the bacterial genome.

@@

% Biological motivations: S. mutans, CCR and CcpA
About one fifth of children of ages ranging from 2 to 19 were reported to have
untreated cavities in the United States \citep{health2010}. Oral bacteria digest
foods containing carbohydrates left on the teeth to convert them to acids, which
dissolve the surface of the teeth and create cavities.  The primary etiological
agent of human dental caries, \textit{Streptococcus mutans}, is one of hundreds
bacterial species that constitute the oral microbial community
\citep{Jenkinson2005}.  Bacteria can selectively use substrates from different
carbon sources via a regulation called carbon catabolite repression (CCR)
\citep{Gorke2008}.  \textit{S.\ mutans} also possesses biological systems for
CCR.  CCR were known to regulates the expression of virulence factors in many
pathogenic bacteria, thereby subjecting a number of genes to CCR.  CcpA was
shown to regulate metabolism and virulence gene expression in 
\textit{S.\ mutans} through CCR in transcriptome study using 
the microarray technique based on DNA hybridization \citep{Abranches2008}.

% Technological motivations: RNA-seq, and microarray, bacterial transcriptomes
It has been a decade since the first genome of the oral bacteria was finished
\citep{Ajdic2002}.  Studies of the transcriptome in the bacteria were performed
based on DNA hybridization
\citep[e.g.,][]{Klein2010,Lemme2011,Redanz2011,Wen2011} whereas those in other
bacterial species took advantage of high-throughput sequencing technologies:
e.g.,
\textit{Mycoplasma pneumoniae} \citep{Guell2009},
\textit{Methanosarcina mazei} \citep{Jaeger2009},
\textit{Bacillus subtilis}  \citep{Irnov2010},
\textit{Helicobacter pylori} \citep{Sharma2010}, and
\textit{Sulfolobus solfataricus} \citep{Wurtzel2010}, to name a few.
The advent of the high-throughput RNA deep sequencing technologies allowed
annotation-free detection of transcripts, sensitive and dynamic ranges of
measurements in RNA expression levels, and studies of non-coding RNAs.

% Aims of the study
We used the standard RNA deep sequencing technique to sequence RNAs in
\textit{S.\ mutans}, focusing on messenger RNAs expression levels in response to
different nutrients with UA159 wild-type and \textit{ccpA}-deleted TW1
mutant-type.  We describe transcripts of the mRNAs using the RNA-seq
data, and present several small non-coding RNAs in \textit{S.\ mutans}. 
Using the map of transcripts
and expression levels of the transcripts from different biological conditions
allowed us to determine which transcripts and non-coding small RNAs were
differentially expressed. This provided more refined views of dynamic changes in
the transcriptome landscape of the etiological agent of human dental caries,
\textit{S.\ mutans}, than previously reported by \citet{Abranches2008}.

@@

<<eval=FALSE,echo=FALSE,results=hide>>=
# not run when smutans R package is installed
# Use this when smutans R package is being tested.
library(DESeq)
library(GenomicRanges)
library(plotrix)
source("smutans/R/class_and_slots.R")
source("smutans/R/core.R")
source("smutans/R/methods.R")
load("smutans/data/smutansData.tx.RData")
load("smutans/data/smutansData.txGenes.RData")
load("smutans/data/smutansData.txGrCombined.RData")
load("smutans/data/smutansData.txAnnotation.RData")
load("smutans/data/smutansData.txRNAz.RData")
load("smutans/data/smutansData.txPileup.RData")
load("smutans/data/smutans.feature.genes.RData")
load("smutans/data/smutans.genes.criteria.RData")
load("smutans/data/smutans.go.genes.RData")
load("smutans/data/smutans.cat.desc.RData")
load("smutans/data/smutans.keg.genes.RData")
load("smutans/data/smutans.keg.desc.RData")
load("smutans/data/smutansGenes.RData")
load("smutans/data/smutansGenes2.RData")
@
<<echo=FALSE,results=hide>>=
# not run without smutans R package installed
# Remove eval=FALSE when the R package is installed
library(smutans)
inDir = system.file("extdata", package="smutans", mustWork=TRUE)
smutansGenes <- 
  readSmutans (countsFile=file.path(inDir, "count-ua159-bwa.txt"),
               indexFile =file.path(inDir, "count-ua159.txt.index"),
               condition = c("UA159GLU", "UA159GAL", "TW1GLU", "TW1GAL"),
               firstFactor = c("UA159GLU", "UA159GAL"),
               firstFactorLabel = c("ua159", "tw1"),
               secondFactor = c("UA159GLU", "TW1GLU"),
               secondFactorLabel = c("glucose","galactose"),
               name="S. mutans UA159, TW1, Glucose, and Galactose")
data(smutansData.tx)
data(smutansData.txGenes)
data(smutansData.txGrCombined)
data(smutansData.txAnnotation)
data(smutansData.txRNAz)
data(smutansData.txPileup)
data(smutans.feature.genes)
data(smutans.genes.criteria)
data(smutans.go.genes)
data(smutans.cat.desc)
data(smutans.keg.genes)
data(smutans.keg.desc)
data(smutansGenes2)
@

<<echo=FALSE,result=hide>>=
notExpressedTx <- smutansData.txGrCombined[elementMetadata(smutansData.txGrCombined)$zero > 0.1 & elementMetadata(smutansData.txGrCombined)$pileup < 5]
grTx <- smutansData.txGrCombined[!(elementMetadata(smutansData.txGrCombined)$zero > 0.1 & elementMetadata(smutansData.txGrCombined)$pileup < 5)]
# We query genes and annotations (genes and RNAs) againt tx's.
matchGenesTx <- findOverlaps(smutansData.txGenes,grTx)
matchAnnotationTx <- findOverlaps(smutansData.txAnnotation,grTx)
# I use gaps with IRanges because gaps with GenomicRanges does not work as expected.
grNoAnnotation <- gaps(ranges(smutansData.txAnnotation))
grNoncoding <- gaps(ranges(smutansData.txGenes))
grRNAz <- ranges(smutansData.txRNAz)
# 
matchUnannotatedRNA <- findOverlaps(grRNAz,grNoAnnotation,type="within")
txRNAz.unannotated <- grRNAz[as.matrix(matchUnannotatedRNA)[,1]]

# A bug in GenomicRanges: the last segment is the while genome. I removed it
# from the noncoding ranges.
# matchNoncoding <- findOverlaps(grNoncoding,grTx)

# Expressed transcripts without annotated genes
grTxWGenes <- grTx[unique(as.matrix(matchGenesTx)[,2])]
grTxWOGenes <- grTx[-unique(as.matrix(matchGenesTx)[,2])]
grTxWOAnnotation <- grTx[-unique(as.matrix(matchAnnotationTx)[,2])]

# Find transcripts with genes with conflicted strands
grGenesPlus <- smutansData.txGenes[strand(smutansData.txGenes)=='+']
grGenesMnus <- smutansData.txGenes[strand(smutansData.txGenes)=='-']
#mtchPlus <- findOverlaps(grGenesPlus,grTxWGenes)
#mtchMnus <- findOverlaps(grGenesMnus,grTxWGenes)
#txWithConflictedGenes <- intersect(as.matrix(mtchPlus)[,2],as.matrix(mtchMnus)[,2])

# Tx with genes strands
grTx2 <- grTxWGenes[!strand(grTxWGenes)=='*']

# Find the number of genes in segments.
matchGenesTx2 <- findOverlaps(smutansData.txGenes,grTx2,type="within")
number.genes.tx <- runLength(Rle(as.matrix(matchGenesTx2)[,"subjectHits"]))

# Length distribution of UTR regions
# grTx2 <- grTxWGenes[-txWithConflictedGenes] 
# Lengths of 5' and 3' UTRs
mtchPlus <- findOverlaps(grGenesPlus,grTx2,type="within")
mtchMnus <- findOverlaps(grGenesMnus,grTx2,type="within")
length5UTR <- c()
length3UTR <- c()
for (i in unique(as.matrix(mtchPlus)[,"subjectHits"])) {
  x <- as.matrix(mtchPlus)[as.matrix(mtchPlus)[,"subjectHits"]==i,"queryHits"]
  y <- min(start(grGenesPlus[x])) - start(grTx2[i])
  z <- end(grTx2[i]) - max(end(grGenesPlus[x])) 
  length5UTR <- c(length5UTR,y)
  length3UTR <- c(length3UTR,z)
}
for (i in unique(as.matrix(mtchMnus)[,"subjectHits"])) {
  x <- as.matrix(mtchMnus)[as.matrix(mtchMnus)[,"subjectHits"]==i,"queryHits"]
  y <- min(start(grGenesMnus[x])) - start(grTx2[i])
  z <- end(grTx2[i]) - max(end(grGenesMnus[x])) 
  length5UTR <- c(length5UTR,z)
  length3UTR <- c(length3UTR,y)
}

# Correlation of expression levels of adjacent genes.
pileup.x <- lapply(ranges(smutansData.txGenes), function(x) seqselect(smutansData.txPileup, x))
pileup.y <- mapply(function(x,y) sum(x)/y, x=pileup.x, y=width(smutansData.txGenes))
elementMetadata(smutansData.txGenes)["score"] <- pileup.y
mtch3 <- findOverlaps(smutansData.txGenes,grTx2)
x <- rle(as.matrix(mtch3)[,"subjectHits"])
comp.exp.two.genes <- c()
for (i in x$values[x$lengths > 1]) {
  y <- combn(as.matrix(mtch3)[,"queryHits"][as.matrix(mtch3)[,"subjectHits"]==i],2)
  for (j in 1:dim(y)[2]) {
    comp.exp.two.genes <- cbind(comp.exp.two.genes,as.matrix(elementMetadata(smutansData.txGenes)["score"][y[,j],]))
  }
}
# Correlation of expression levels of adjacent genes between transcripts.
comp.exp.two.genes.between.transcript <- c()
for (i in 1:length(x$values)) {
  if (i == length(x$values)) {
    break
  }
  y1 <- as.matrix(mtch3)[,"queryHits"][as.matrix(mtch3)[,"subjectHits"]==x$values[i]]
  i2 <- i + 1
  y2 <- as.matrix(mtch3)[,"queryHits"][as.matrix(mtch3)[,"subjectHits"]==x$values[i2]]
  l <- as.matrix(do.call(expand.grid, list(y1,y2)))
  for (j in 1:dim(l)[1]) {
    comp.exp.two.genes.between.transcript <- cbind(comp.exp.two.genes.between.transcript,as.matrix(elementMetadata(smutansData.txGenes)["score"][l[j,],]))
  }
}
@

<<eval=FALSE,echo=FALSE,result=hide>>=
# To find Infernal small RNAs expressed in RNA-seq data
grTx <- smutansData.txGrCombined[!(elementMetadata(smutansData.txGrCombined)$zero > 0.1 & elementMetadata(smutansData.txGrCombined)$pileup < 5)]
smutansData.txInfernal <- smutansData.txAnnotation[1:274]
ol.Infernal <- findOverlaps(smutansData.txInfernal,grTx,type="within")
r <- ranges(grTx[unique(ol.Infernal@subjectHits)])
for (i in seq(length(r))) {
  cat(paste("chr1:",start(r)[i],"-",end(r)[i]),"\n")
}
#chr1: 22518 - 23227 tRNA
#chr1: 23229 - 25062 tRNA
#chr1: 97029 - 99661 mir
#chr1: 99934 - 100544 tRNA
#chr1: 111689 - 118470 mir
#chr1: 152804 - 153769 * STnc490k different position
#chr1: 182583 - 185313 mir
#chr1: 190870 - 191839 tRNA
#chr1: 293153 - 293660 * SRP_bact SRP_euk_arch different position
#chr1: 322211 - 328546 tRNA
#chr1: 413043 - 415691 * T-box
#chr1: 552584 - 553935 mir
#chr1: 626450 - 630612 * P15 (ambiguous) 
#chr1: 660501 - 661809 * L20_leader (strong) LR-PK1 (ambiguous)
#chr1: 666062 - 667070 * TPP (ambiguous but strong)
#chr1: 710297 - 712890 mir
#chr1: 796924 - 798060 * L21_leader (ambiguous)
#chr1: 800164 - 814296 * PyrR (ambiguous)
#chr1: 938383 - 938619 mir
#chr1: 995160 - 998898 snoR72 (ambiguous)
#chr1: 1115782 - 1117618 * Glycin (ambiguous)
#chr1: 1140867 - 1142427 tRNA
#chr1: 1142434 - 1149502 tRNA
#chr1: 1163587 - 1166184 PyrR (ambiguous)
#chr1: 1209827 - 1214087 tRNA
#chr1: 1214631 - 1215039 * L19_leader (ambiguous)
#chr1: 1326825 - 1328325 * CRISPER (ambiguous)
#chr1: 1410520 - 1420619 * His_leader (ambiguous)
#chr1: 1439314 - 1447074 * T-box (ambigous)
#chr1: 1447225 - 1451227 Thr_leader
#chr1: 1487846 - 1488056 mir
#chr1: 1605925 - 1606244 mir
#chr1: 1615311 - 1618362 23S-methyl
#chr1: 1654702 - 1655118 tRNA
#chr1: 1677566 - 1678101 * T-box (ambiguous)
#chr1: 1678104 - 1678469 tRNA
#chr1: 1758499 - 1759936 mir
#chr1: 1783980 - 1787743 mir
#chr1: 1838690 - 1844933 * mir
#chr1: 2023401 - 2026533 tRNA mir

# To find RNAz predicted small RNAs expressed in RNA-seq data
ol <- findOverlaps(smutansData.txRNAz,grTx,type="within")
r <- ranges(grTx[unique(ol@subjectHits)])
for (i in seq(length(r))) {
  cat(paste("chr1:",start(r)[i],"-",end(r)[i]),"\n")
}
#chr1: 301108 - 303712 no
#chr1: 303714 - 304038 * locus100 His_leader (weak)
#chr1: 340223 - 341140 * locus101 
#chr1: 378392 - 382328 
#chr1: 413043 - 415691 * locus103 T-box (strong)
#chr1: 480819 - 482162 * locus104 (ambiguous)
#chr1: 681862 - 684666 * locus107 yybP-ykoY (ambiguous)
#chr1: 742326 - 747734 * locus108 (ambiguous)
#chr1: 908974 - 911426 * locus110 (weak)
#chr1: 1036181 - 1038689 locus2
#chr1: 1209827 - 1214087 locus4 tRNA (false?)
#chr1: 1304269 - 1305793 locus7 (ambiguous)
#chr1: 1678104 - 1678469 locus53 tRNA
#chr1: 1783980 - 1787743 locus69

# Expressed regions without annotations
smutansData.txAllAnnotation <- c(smutansData.txRNAz,smutansData.txAnnotation)
strand(smutansData.txAllAnnotation) <- '*'
smutansData.txAllAnnotation <- gaps(reduce(smutansData.txAllAnnotation))
smutansData.txAllAnnotation <- smutansData.txAllAnnotation[-2:-1]
# Check sum(width(smutansData.txAllAnnotation)) is small or < 10% of 2Mbp.
x <- c()
for (i in seq(length(smutansData.txAllAnnotation))) {
  s <- start(ranges(smutansData.txAllAnnotation))[i]
  e <- end(ranges(smutansData.txAllAnnotation))[i]
  x <- c(x,sum(smutansData.txPileup[s:e])/(e-s+1))
}
elementMetadata(smutansData.txAllAnnotation)[,"score"]<-x
smutansData.txNoAnnotation <- smutansData.txAllAnnotation[order(elementMetadata(smutansData.txAllAnnotation)[,"score"])]
for (i in seq(length(smutansData.txNoAnnotation))) {
  if (abs(start(smutansData.txNoAnnotation)[i] - end(smutansData.txNoAnnotation)[i]) > 50) {
    cat(paste("chr1:",start(smutansData.txNoAnnotation)[i],"-",end(smutansData.txNoAnnotation)[i]),"\n")
  }
}
# This does not work. I think that I have to manually check.
strand(smutansData.txNoAnnotation) <- '+'
x <- elementMetadata(smutansData.txNoAnnotation)[,"score"]
x <- as.integer(x*1000/max(x))
elementMetadata(smutansData.txNoAnnotation)[,"score"] <- x
smutansData.txNoAnnotation <- sort(smutansData.txNoAnnotation)
elementMetadata(smutansData.txNoAnnotation)[,"name"] <- seq(length(smutansData.txNoAnnotation))
export.bed(smutansData.txNoAnnotation,con="b.bed")
x <- c(46, 69, 73, 74, 75, 76, 80, 97, 103, 212, 261, 280, 288, 317, 440, 450, 478,
484 , 628 , 753 , 1011, 1019 , 1037 , 1084, 1097, 1176, 1291, 1444, 1508,
1518, 1600, 1605, 1638, 1645, 1690)
for (i in x) {
  cat(paste("chr1:",start(smutansData.txNoAnnotation)[i],"-",end(smutansData.txNoAnnotation)[i]),"\n")
}

# 46
# 69 * strong
# 73 ambiguous
# 74 very weak
# 75 ambiguous
# 76 ambiguous
# 80 * weak
# 97 *
# 103 *
# 212 * strong
# 261 *
# 280 *
# 288 * one gene?
# 317 * in between tRNAs
# 440 weak
# 450 * strong
# 478 * strong
# 484 * strong
# 628 *
# 753 *
# 1011 * strong
# 1019 * tail of a gene?
# 1037 * strong?
# 1084 * strong
# 1097 * left? PCR amplification error?
# 1176 * start or end of genes?
# 1291 * end of the gene?
# 1444 * 
# 1508 * strong
# 1518 * strong
# 1600 * start of the gene?
# 1605 * new gene in between genes?
# 1638 * start of the gene?
# 1645 * start of the gene?
# 1690 * in between genes?

for (i in seq(length(txRNAz.unannotated))) {
  cat(paste("# ",names(txRNAz.unannotated)[i],"  chr1:",start(txRNAz.unannotated)[i],"-",end(txRNAz.unannotated)[i]),"\n")
}
#  locus99   chr1: 302288 - 302403  no
#  locus100   chr1: 303745 - 303859 did
#  locus101   chr1: 340263 - 340343 did
#  locus102   chr1: 378452 - 378505 good one
#  locus104   chr1: 481458 - 481539 not a good candidate
#  locus106   chr1: 674067 - 674135 maybe one
#  locus108   chr1: 747483 - 747567 not a good one
#  locus110   chr1: 911157 - 911258 not a good one
#  locus2   chr1: 1036231 - 1036283 not a good one
#  locus5   chr1: 1244007 - 1244086 not a good one
#  locus6   chr1: 1297150 - 1297227 not a good one
#  locus7   chr1: 1304792 - 1304926 not a good one
#  locus8   chr1: 1313525 - 1313674 not a good one
#  locus10   chr1: 1564032 - 1564149 not a good one
#  locus69   chr1: 1787575 - 1787687 not a good one
#  locus85   chr1: 1864317 - 1864461 not a good one
@

\subsection*{Small RNAs expressed in RNA-seq}

We compiled 274 predicted RNAs by using cmsearch program from Infernal (v1.0)
with default options for models in RFAM 9.1 \citep{Gardner2009,Nawrocki2009}.
By intersecting the predicted RNAs locations and RNA-seq expressed regions,
we found more evidence for some of the predicted RNAs (table \ref{tab:infernal}).
Several RNAs were shown to be clearly expressed by RNA-seq experiments:
T-box strong, RNaseP\_bact\_a, RNaseP\_bact\_b, L20\_leader, tmRNA, FMN, and 6S.
Other RNAs, TPP and Glycin, were located close to their neighboring genes,
making it hard to discriminate them from the genes.

Additionally, we predicted 
\Sexpr{length(smutansData.txRNAz)}
structural RNAs using RNAz, most of which were overlapped with those from
the Infernal prediction. We found a few of them to be expressed based on RNA-seq
experiments (table \ref{tab:RNAz}).
Region locus100 had tens of short reads mapped on it, which was near
His\_leader predicted by Infernal. 
Although region locus101 was ambiguous in its location with respect to its neighboring
genes, its predicted secondary structure hinted at its potential as a functional
RNA. Region locus103 was expressed clearly compared with neighboring regions; the
region was predicted as T-box by Infernal.
Although region locus107 was ambiguous in its location with respect to its neighboring
genes expression, its good secondary structure and yybP-ykoY predicted by
Infernal hinted at the RNA functional potential of the region.

\subsection*{Unannotated regions expressed in RNA-seq}

We focused 16 regions that had not been annotated either by previously known
genes or by the potentially functional RNA prediction of our current study
(table \ref{tab:unannotated}). Four regions inlcuding novel6,
novel7, novel10, and novel11 showed that their expression levels were not
clearly dicriminated from their flanking genes. 
Located in between SMU.305 and SMU.307 was, for example, novel6, which was
connected in expression level by large number of short reads with SMU.307.
The other 12 regions appeared to show distinct expression levels compared with
flank regions. Region novel1 was situated in between SMU.61 and SMU.63c, which
had different strands.
Although seemingly separated from either of the two genes, it was connected by
links of average about 20 depths of short reads with the either of them.
Regions novel2, novel13, and novel15 appeared to be expressed very weakly.
A few regions were neighboring known RNAs: 
novel3 located in between SMU.97 and SMU.t19,
novel4 located in between SMU.t20 and SMU.107, and
novel6 situated to the right of
SRP\_bact and SRP\_euk\_arch predicted by Infernal.
We also examined regions in between known CDS or tRNAs, which could be
unannotated regions but likely to be in operon with its neighboring genes
(table \ref{tab:inbewteencds}).

% Of the \Sexpr{length(colnames(smutans.genes.criteria))-1} predicted non-coding RNAs
% \Sexpr{length(txRNAz.unannotated)} were in  
% unannotated non-coding regions (table \ref{tabUnannotatedRNA}).

Using
\program{RNAplex} \citep{Tafer2008} and \program{RNAplfold} \citep{Bernhart2006}
we predicted target genes of the putative non-coding RNAs.  Table \ref{tab:goseq}
shows functional categories enriched with target genes for predicted small RNAs.

We found that a few of the putative small RNAs were structured with stable stems
(figs.\ \ref{figRNALocus1}, \ref{figRNALocus2}, and \ref{figRNALocus3}).

\subsection*{Predicted transcripts}

We obtained a set of segments of RNA-seq expressions for each of the 13 RNA-seq
samples.  All of the segments for each set of RNA-seq expression segments were
scored by the average proportions of segments shared by the segments in the rest
of sets.  We placed more highly ranked segments first on the reference genome
based on the scores, and purged more lowly ranked segments from the
transcriptome map. In doing so, the
final set of segments was formed by placing non-overlapping highly ranked
segments.  Average site-wise expression levels of the segments 
are shown at figure \ref{figGrCombinedExpression}. 
We designated segments as expressed that had site-wise pileup values
greater than 5 and proportion of sites of zero site-wise pileup values less than
10\%.  A gene was also designated as an expressed if the gene belonged to an
expressed segment.  Some of the predicted segments were not expressed or very
lowly expressed (\Sexpr{length(notExpressedTx)} 
out of 
\Sexpr{length(smutansData.txGrCombined)}).
Of the total \Sexpr{length(smutansData.txGenes)} of genes,
\Sexpr{length(as.matrix(matchGenesTx)[,1])}
(\Sexpr{as.integer(length(as.matrix(matchGenesTx)[,1])/length(smutansData.txGenes)*100)}\%) 
were designated as expressed. 
Genes that were not designated as expressed could be due to false gene
annotations or possibility of true negative expression under the experimental
conditions. 
Figure \ref{figLength} shows the length distribution of expressed and
not-expressed genes. 
Of the \Sexpr{length(grTx)} expressed segments, 
\Sexpr{length(grTxWOAnnotation)} contained no annotated genes or RNAs.
Many of the segments might cover gaps between real operons.
\fixme{Examine these unannotated regions in the UCSC genome tracks.}

Figure \ref{figTranscriptLength} shows the length distribution of expressed
segments with annotated coding genes and unannotated regions.  Unannotated
regions could include potential non-coding regions and novel coding genes.
Segments of unannotated regions appeared to be shorter than those with
coding genes. Most of the unannotated regions with length being near 360 bp
could be non-coding regions, and the rest of the unannotated regions with length
being near 1000 bp could be new coding genes. Some of the unannotated genes
could be pseudogene regions.  We used strandnesses of known genes to find
strands of segments although we could not tell strands of the predicted
transcripts solely based on our RNA-seq data because of our strand-nonspecific
RNA-seq technique.  Of
the \Sexpr{length(grTx)} expressed transcripts, 
\Sexpr{sum(strand(grTx)=='*')}
included annotated genes of both strands
(\Sexpr{as.integer(sum(strand(grTx)=='*')/length(grTx)*100)}\%).
We filtered out the transcripts with genes with conflicted both strands leaving
\Sexpr{length(grTx2)} expressed transcripts. 
Figure \ref{figUTRLength} shows length distributions of 5' and 3' UTR of the
expressed transcripts without genes with conflicted strands.  Many transcripts
appeared to start and end near the boundaries of annotated genes.  Gene
expression levels for adjacent genes within expressed transcripts were more
correlated than those between adjacent transcripts 
(fig.\ \ref{figExpressionTwoGene}).
Among the \Sexpr{length(unique(as.matrix(matchGenesTx2)[,"subjectHits"]))} segments
with genes of all of the same strands, 
\Sexpr{sum(number.genes.tx==1)} segments contained a single gene while 
\Sexpr{sum(number.genes.tx > 1)} contained more than one gene.

The automatically defined transcripts using site-wise expression values by the
hidden Markov model and mapped short reads displayed using UCSC genome browser
tracks guided us to define transcripts.

tx0391
tx0392

@@

First using the table of read counts of annotated genes we formed a cluster of
the 13 RNA-seq samples (fig.\ \ref{figClust}). The effect of nutrients appeared
to be stronger than that of the deletion of \textit{ccpA} genes from the UA159
genome.

<<echo=FALSE,result=hide>>=
glucoseOnly <- newSmutans( smutansGenes, title="Glucose Only" )
glucoseOnly <- smutans.de2( glucoseOnly, condition="glucose", 
                            condA="ua159", condB="tw1" )
save(glucoseOnly, file=file.path("glucoseOnly.RData"))
#load("glucoseOnly.RData")

smutans.de2List( glucoseOnly, "glucoseOnly.csv" )
tableGlucoseOnly <- smutans.de2List( glucoseOnly, qval=1e-3 )

# Gene descriptions
geneDescription <- read.delim("feature-genome.out-geneonly.txt",head=FALSE,sep="\t")
geneDescription <- geneDescription[,c(4,7)]
colnames(geneDescription) <- c("Gene ID","Gene Description")
# Functional classes
functionalDescription <- read.csv("feature-genome.out-geneonly-oralgen-functional.txt",head=FALSE)
functionalDescription <- functionalDescription[,c(2,12)]
colnames(functionalDescription) <- c("Gene ID","Functional Class")
# Find potential differentially expressed genes, and add gene descriptions, and
# functional classes
x <- smutans.de2List( glucoseOnly, qval=1e-3 )
x <- x[,c("id","baseMeanA","baseMeanB","log2FoldChange","padj")]
colnames(x) <- c("Gene ID","UA159","TW1","fold", "p-value")
x <- merge(x, geneDescription, by.x = "Gene ID", by.y = "Gene ID")
x <- merge(x, functionalDescription, by.x = "Gene ID", by.y = "Gene ID")
tableGlucoseOnly <- x
@

Two independent factors of the biological processes included the effect of CcpA
and the treatment of mono-saccharides (glucose versus galactose), resulting in
four pair-wise comparisons. We compared two conditions of one factor with the
other factor being fixed at one condition. For example, we compared UA159
wild-type and TW1 mutant-type strains under the condition of glucose-grown cells
to determine differentially expressed transcripts or genes potentially due to
presence and absence of deactivation of \textit{ccpA} genes.  We performed a
similar comparison with galactose-grown cells. We replicated four
biological experiments for the condition combination of UA159 wild-type grown
in galactose, and three replicates for the other three conditions.

\subsection*{Comparison between wild-type and mutant-type in glucose-grown cells}

Table \ref{tabGlucoseOnly} lists differentially expressed genes in the
comparison of UA159 wild-type and TW1 mutant-type grown with glucose.
Of \Sexpr{nrow(counts(smutansGenes))} genes, 
\Sexpr{nrow(tableGlucoseOnly)} genes were found to be differentially expressed
at FDR of 0.1\%.  We confirmed that the most up-regulated genes in the TW1
mutant-type encoded the components of the PDH enzyme complex (SMU.1421,
SMU.1422, SMU.1423, and SMU.1424) as shown previously.  
These genes appeared to be transcribed in
a single operon (fig.\ \ref{figUCSC}). 
Figure \ref{figPieGlucoseOnly} shows the
proportions of \program{oralgen} databases functional classes by the number of
differentially expressed genes.  Notably, we found
\Sexpr{length(grep("Energy metabolism",tableGlucoseOnly[,"Functional Class"]))}
genes that were involved in energy metabolism where sugars and fermentation
subclasses were profound. We found 
\Sexpr{length(grep("PTS",tableGlucoseOnly[,"Functional Class"]))}
genes in PTS class of signal transduction. Another subclass, 
``carbohydrates, organic alcohols, and acids,'' in
``transport and binding proteins'' was following next in the number of
differentially expressed genes. We also found  
\Sexpr{length(grep("hypothetical",tableGlucoseOnly[,"Gene Description"]))}
genes annotated as hypothetical, which might constitute the unknown functional
class. These results appeared to be consistent with those of \citet{Abranches2008}.

>>>>

\documentclass{article}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{color} 
\usepackage{setspace} 
\usepackage{rotating}
\usepackage{natbib}
\usepackage{xr}

@@

\makeatother
\pagestyle{myheadings}
\markright{Transcriptome of Streptococcus agalactiae}

@@

\title{Defining Transcriptome using RNA Sequencing\\
of {\em Streptococcus agalactiae} }

\author{Vince P.\ Richards$^{1}$, Sang Chul Choi$^{2}$,
and Adam Siepel$^{2}$, and Michael J.\ Stanhope$^{1}$}

@@

\begin{footnotesize}
\begin{center}
$^1$Department of Population Medicine and Diagnostic Sciences,\\
College of Veterinary Medicine, Cornell University, Ithaca, NY 14853, USA
\\[1ex]
$^2$Department of Biological Statistics and Computational Biology,\\
Cornell University, Ithaca, NY 14853, USA
\end{center}
\end{footnotesize}

@@

\begin{tabular}{lp{4.5in}}
{\bf Submission type:}& Research Article
\vspace{1ex}\\
{\bf Keywords:}&RNA-seq, non-coding RNAs
\vspace{1ex}\\
{\bf Running Head:}&Transcriptome of {\em Streptococcus agalactiae}
\vspace{1ex}\\ 
{\bf Corresponding Author:}&
\begin{minipage}[t]{4in}
 Adam Siepel\\
 102E Weill Hall, Cornell University\\
 Ithaca, NY 14853\\
 Phone: +1-607-254-1157\\
 Fax: +1-607-255-4698\\
 Email: acs4@cornell.edu
\end{minipage}
\end{tabular}

@@

\setcounter{page}{1}

@@

A bacterial strain NEM316 of {\em Streptococcus agalactiae} has been known to
originate in humans. We provide a series of evidence for the hypothesis that it
is originated from cows.

@@

@@

@
<<echo=FALSE,results=hide>>=
library(smutans)

for (i in c("NC_004368")) {
file.1 <- paste("../output/agalactiae/1/bwa/count-",i,".txt", sep="")              
file.out.1 <- paste(i,"-sag56-A-milk-B-broth-clust.pdf", sep="")                         
file.out.2 <- paste(i,"-sag56-A-milk-B-broth.csv", sep="")                               
file.out.3 <- paste(i,"-sag56-A-milk-B-broth.ps", sep="")

agGenes <-
  readSmutans(countsFile=file.1,
              indexFile="../output/agalactiae/1/run-analysis/count.txt.index",
              condition = c("sag56hm", "sag56hb"),
              firstFactorLabel = c("NEM316"),
              secondFactorLabel = c("sag56hm", "sag56hb"),
              name="S. agalactiae NEM316 in milk versus broth")

ag <- newSmutans( agGenes , title="S. agalactiae NEM316 in milk versus broth")

pdf(file.out.1)                                                                 
smutans.de2Clust( ag )                                                          
dev.off()

ag <- smutans.de2( ag, type="NEM316", condA="sag56hm", condB="sag56hb")
smutans.de2List( ag, file=file.out.2 )                                          
smutans.plotDiffExp(ag,file=file.out.3,qval=0.05)                               
}
@

@@
